A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-06 natural history of scavenging.tex
Edits up to handling time part

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals.
This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour \citep{devault2003scavenging}.
Professor Sanborn Tenney writing in 1877 for The American Naturalist had this to say about one well known group, ``prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters."
He contrasts these with ``nobler kinds" of carnivores such as lions and tigers \citep{tenney1877few}.
Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact.
Without catching a carnivore in the act of killing we are left to infer how the prey was killed.
Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.
But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted.
The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms.
Indeed, a single species of dinosaur notwithstanding \citep{carbone2011intra}, a synthesis describing the natural history of scavengers is absent from the literature.
%TG: I don't understand the dinosaur part of this sentence. Might be my non-native english but I thought it's worth noting for other non-english folk out there (even though I know they aren't many!).
Fortunately, research on scavenging is on the rise \citep{manga2006vulture}.
As a result, we are now beginning to realise the extent of this behaviour such that, ``in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{beasley2015vertebrates}.
This has profound implications for the trophic ecology of these systems and particuarly our models of them.
Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses \citep{jones2015african}.
By recognising the difficulty in directly observing scavenging, other methods have been turned to to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper.
Here we chart the natural history of scavenging by assessing the potential for the behaviour in dominant vertebrate groups given their ecology and functional traits.
%It is our aim in this review to employ these methods to gain an understanding of scavengers past and present.

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Thus, any animal existing as a scavenger must minimise its locomotory costs and maximise its detection capabilities \citep{ruxton2004obligate}.
Once found, the scavenger has to process the carrion and overcome the agents of decay produced by the action of microorganisms on the carcass in order to take in a suffienct amount of energy \citep{ruxton2014fruit}. 
The habitat must also be productive enough to sustain a pool of animal biomass that will eventually produce carcasses.
These facets are essentially the key parameters found in the species functional response curves of optimal foraging theory, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. 
Here, we can draw on these aspects to envisage a scavenger moving through its environment, searching for food and trying to process it efficiently as we explore the prevalence of scavenging through time. 
Our aim is to produce a 'scale of scavenging' whereupon we can place any species, past or present, and assess the importance of scavenging in its diet. 
% The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
%TG: Insert figure here: scavenging scale + four points: Locomotion/Detection/Consumption/Environment
\section*{Encounter Rate}

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts %TG: not sure about "bouts" get it with the context and I know it makes the whole review a bit more literary and thus nicer to read but it might exclude non-native english readers. 
 of energetically expensive activity \citep{bennett1979endothermy}. 
By contrast, modern reptiles are ectothermic, limiting their activity periods \citep{bennett1979endothermy}.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in the phenomenon known as Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
These constraints manifest in aspects such as maximum sustainable speed where an equivalent mammal enjoys a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, %TG: or maybe no brackets... Up to you Adam.
 are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
We can (and have) use(d) these insights to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Million years ago; Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya) were likely to have been ambush predators which would argue against them taking a lot of carrion.  
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

Of course, the importance of terrestrial tetrapods predates the evolution of the dinosaurs. %TG: I will down that "dominance" term. Maybe change the sentence to "Of course the importance of terrestrial tetrapods in ecosystems predates the evolution of the dinosaurs." AK Done 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
And it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Scavenging behaviour may have evolved on land as soon as the first terrestrial tetrapods emerged.
In fact, some of the earlier tetrapods tracks dating back to the early Middle Devonian (393.3 - 387.7 Mya) were found in intertidal environments \citep{Niedzwiedzki2009}.
These environments are isolated from marine systems twice a day leaving potential carrion unexploited by marine vertebrates.
\cite{Niedzwiedzki2009} suggest that these environments ``would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource.''

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers \textit{par excellence}. %TG: not sure how reviewers will go with French but I like it!
Certainly, scavenging should be particularly attractive to avian predators compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still for searching the environment \citep[approximately twice the basal metabolic rate; ][]{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
The difficulty inherent in observing marine birds foraging at sea means our knowledge of their feeding habits is limited. 
But it does seem that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}.
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}. 
In vultures, who are more easily observed on land, we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers to date. %TG: Or "obligate vertebrate scavengers known to date." For a more emphasizing turn.

%TG: I thought the paragraph was a bit long so here's a cut for "flying scavengers in the past"
The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica}, so many of the benefits of flight would have been realised from that point on for carnivorous birds.
Vertebrate flight is much older still, where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.
However, some fossil from the Eocene (56 to 33.9 Mya) suggest the presence of possible scavenging bats.
A large bat from the middle to late Eocene (56 to 33.9 Mya) with a robust cranio-mandibular morphology, \textit{Necromatis} (``death-eater''), could have intake carrion as a significant proportion of its diet \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.
Although, this example might be peculiar to this exceptional fossil given most other remains represent animals less suited to the lifestyle \citep{eiting2009global}.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution}. 
This has led some researchers to argue for the feasibility of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
As in aerial and terrestrial environments we have evidence of facultative scavenging among extinct aquatic species.
For example, shark teeth have been found in the remains of dinosaurs, mososaurs and 4-million-year-old mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.  
%the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark, \textit{Squalicorax} \citep{schwimmer1997scavenging}.
%As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and  whale from Peru \citep{ehret2009caught}.
Extant White sharks (\textit{Carcharodon carcharias}) too are known to feed on whale carcasses \citep{fallows2013white}.

We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers. 
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, their total energetic cost is similar to an equivalent terrestrial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

@@

\subsection*{Detection}
It would be pointless to have incredible ranging abilites and not have the sensory architecture to benefit from it.
If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed, this is what we find. 
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
If we compare this to the energetics approach by \cite{ruxton2004obligate}, they calculated a terrestrial scavenger needs to be able to detect carrion at 500 m in order to survive, which is clearly within the ability of hyenas.  
% While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
Moreover, the senses of many extant (and in all probabilty extinct) carnivores meet this required distance, making scavenging feasible for terrestrial species \citep{farlow1994speculations,mech2010wolves}.

Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Vultures are known to have impressive visual acuity, with one estimate indicating Lappet-faced Vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed visual abilities \citep{reymond1985spatial}.
It follows from this that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.
We can contrast this ability with bats, whose visual acuity is famously poor. 
It also appears that echolocation would not lend itself to discovering immobile carrion.
Their small size and poor terrestrial ability would also count against them at a carcass \citep{riskin2006terrestrial}.

>>>>

\section*{Introduction}

Historically, scavengers have not been viewed as the most charismatic of animals.
This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour \citep{devault2003scavenging}.
Professor Sanborn Tenney writing in 1877 for The American Naturalist had this to say about one well known group, ``prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters."
He contrasts these with ``nobler kinds" of carnivores such as lions and tigers \citep{tenney1877few}.
Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact.
Without catching a carnivore in the act of killing we are left to infer how the prey was killed.
Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.
But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted.
The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms.
Indeed, a single species of dinosaur notwithstanding \citep{carbone2011intra}, a synthesis describing the natural history of scavengers is absent from the literature.
%TG: I don't understand the dinosaur part of this sentence. Might be my non-native english but I thought it's worth noting for other non-english folk out there (even though I know they aren't many!).
Fortunately, research on scavenging is on the rise \citep{manga2006vulture}.
As a result, we are now beginning to realise the extent of this behaviour such that, ``in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{beasley2015vertebrates}.
This has profound implications for the trophic ecology of these systems and particuarly our models of them.
Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses \citep{jones2015african}.
By recognising the difficulty in directly observing scavenging, other methods have been turned to to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper.
Here we chart the natural history of scavenging by assessing the potential for the behaviour in dominant vertebrate groups given their ecology and functional traits.
%It is our aim in this review to employ these methods to gain an understanding of scavengers past and present.

%the intro needs to lead towards a aim a bit more strongly

\subsection*{The Difficulty of Scavenging}
The chief hurdle of scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some level of opportunistic scavenging chances. %TG: Two times "opportunities", replaced by "chances"
However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet their energetic demands.
%Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. %TG: make the following sentence more sense?
Once found, the scavenger must be able to out-compete any potential competitors and process the carrion that is undergoing decay and thus contains toxins and diseases produced by micro-organisms \citep{ruxton2014fruit}.
Finally, the level of scavenging opportunities available will also depend on the density, size, and quality of carcases produced, all of which are dependent on complex ecosystem dynamics.
These facets which are required for scavenging are essentially the key parameters found in the species functional response curves of optimal foraging theory, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. %TG: Jaysus, this sentence doesn't read well! "Functional response curves of optimal foraging theory" can probably be phrased in a nicer/easier way...
We compare the ability of scavenging across vertebrates focusing on each of these facets and their relationship with scavenging.
By considering scavenging in a framework of optimal foraging and identifying the prerequisite attributes and processes required for scavenging we propose a ``scale of scavenging'' whereupon we can place any vertebrate species, past or present, and assess the importance of scavenging in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between the consumer and resources. 
In the simplest case, this rate can be thought of in terms of a gas diffusion model where the movement of two agents depends only on their relative speed.
As carcasses are stationary, the relative speed between a scavenger and carrion is only dependent on the movement of the scavenger. 
Vertebrates do more than simply bump into resources like gas molecules though, because they can actively detect them through their sensory abilities. 
Therefore, scavenging will also depend on search rates, which are dependent on both species physiology and the dimensionality of the environment \citep{pawar2012dimensionality}.

\subsection*{Locomotion}
In contrast to the importance of locomotion in predation, locomotion relating to scavenging depends more on the ability to efficiently move over large areas. 
This ability generally requires an efficient transfer of metabolic energy into movement which relies on both physiology (i.e. metabolism) and the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
This is because, although powered flight is energetically expensive, species like vultures can exploit air currents using large wingspans which allows them to soar while only doubling their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}. %TG: I didn't get the meaning of this sentence, is soaring doubling the metabolic rate? or is it a word misplacement ("while only doubling their metabolic rate")? and if so, how does that make the point?
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
This mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
Thus, the use of soaring in extinct species as a means for scavenging is also likely.

While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate} powered flight is still an efficient means of locomotion. 
Avian flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic} 
Even taking account of maintenance costs this still bears out, where total cost of movement (J kg\textsuperscript{-1} m\textsuperscript{-1}) scales according to 5.2 $\times$ body mass (kg) \textsuperscript{-0.23} for fliers and 10.7 $\times$ body mass (kg) \textsuperscript{-0.32} for runners \citep{williams1999evolution}.

We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. 
Similarly we would expect that extinct species would also scavenge in a similar fashion depending on the efficiency of their flight. 
For example, early birds such as \textit{Archaeopteryx} are predicted to have been poor, relatively inefficient fliers \citep{nudds2010narrow}. 
%Similarly something about the efficiency of pterosaurs flight. TG: Was this a comment? By the way, about pterosaurs (though it might not be necessary to mention it), most of them where tiny little feckers like Archaeopteryx or even smaller.
The importance of efficient flying over large areas may explain the lack of scavenging behaviour in bats as they are generally nocturnal, a time when they would receive no aid from convective air currents \citep{norberg2012vertebrate}. 
%smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), %AK Don't think it is the case that bats are less efficient than birds in terms of flight 
That said, \textit{Necromantis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) had a robust cranio-mandibular morphology, and is a likely candidate for scavenging behaviour \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.% amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}. %TG: we can cut this number of bat species to avoid getting into stamp collecting debate. + there are effectively a two shit load units of them which makes the Necromantis example not really convincing.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the likelihood of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show very large ranges (e.g. great white sharks (\textit{Carcharodon carcharias})) and have the lowest cost of swimming.%TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
Facultative scavenging is seen in many marine groups including species of extant sharks such as great white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}.
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and mysticete whales from the Pliocene \citep[5.3 to 3.6 Mya;]{schwimmer1997scavenging,ehret2009caught}.

% KH Other types of swimming in fish, other species propelled by fins (turtles and plesiosaurs etc)

% KH snakes - ie undulating

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya %TG: that early???
have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, because an animal must trade off the cost of supporting its body off the ground while propelling itself forward. 
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often mentioned \citep[e.g. benefits in terms of speed, efficiency, muscle effort and manoeuvrability; ]{sullivan2015posture}. 
Yet, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Although, metabolic rate unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also be true of extinct species. 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. 
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizeable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
We can (and have) use(d) these insights to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Million years ago; Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya) were likely to have been ambush predators which would argue against them taking a lot of carrion. 
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...
The evolution of bipedal movement freed up the forelimbs for other purposes e.g. wings, tool-use etc. 
In terms of its cost \cite{alexander2004bipedal} shows that, in the case of humans, we are more economical than predicted while walking and less so while running.

%  KH Something more a bipedal movement.
Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

% KH Something about humans (also the carnivours kangaroo which might be the most efficient mover.)

\subsection*{Detection}
It would be pointless to have incredible ranging abilities and not have the sensory architecture to benefit from it.
%If we came at this from a position of complete ignorance we would predict scavengers to have well-developed senses and indeed, this is what we find. %TG: I don't really like this sentence. How about:
As predicted by the necessity of an increased encounter rate, scavengers have well-developed detection senses.
A simplification of terrestrial, vertebrate scavengers in sensory terms is one of them existing in a two-dimensional plane while foraging for carrion directly.
They can detect carcasses at a range that is defined by the radius of their sensory organs. %, usually the visual and olfactory senses.
As a consequence, they have a much more restricted view of the landscape than do aerial foragers.
Hyenas make up for this in their ability to smell a rotting carcass 4 km away and to hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. 
% Using the approach of \cite{spiegel2013factors} we estimate a spotted hyena %TG: do we need a (\textit{Crocuta crocuta}) here?
%could resolve a 2 metre target at 1 km distance. 
If we compare this to the energetics approach by \cite{ruxton2004obligate}, they calculated a terrestrial scavenger needs to be able to detect carrion at 500 meter in order to survive, which is clearly within the ability of hyenas.  
% While considering prehistoric habitats \cite{ruxton2004obligate} calculated that ``a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m".
Moreover, the senses of many extant (and in all probability extinct) carnivores meet this required distance, making scavenging feasible for terrestrial species \citep{farlow1994speculations,mech2010wolves}.
